EGGs are adhesive!

### 2025-03-14 calco/main.tex
intro

\vspace{.1cm}
\noindent
\begin{minipage}[l]{.78\linewidth}In short, following the double-pushout (DPO) approach
to graph transformation~\cite{xxx}, a rule is given by two arrows $l: K \to L$ and $r: K \rightarrow R$
and its execution requires a match $m: L \to G$ such that the rewriting step from $G$
to $H$ is given by the diagram aside, whose square are pushouts.
  \end{minipage}%
    \hfill
  \begin{minipage}[r]{.20\linewidth }
    \xymatrix@C=.5cm@R=.5cm{
      L \ar[d]_{m}
      & K \ar[r]^r \ar[l]_{l} \ar[d] & R \ar [d] \\
      G & C \ar[r] \ar[l]                    & H
    }
%    \xymatrix@C=.5cm@R=.5cm{
%      L \ar[d]_{m} \ar@{}[dr]|{(1)}
%      & I \ar[r]^r \ar@{>->}[l]_{l} \ar[d] \ar@{}[dr]|{(2)} & R \ar [d] \\
%      G & C \ar@{->}[r] \ar@{>->}[l]                    & H\\
 %     & J \ar@{->}[ul] \ar@{->}[u]|{k} \ar@{->}[ur]
 %   }
  \end{minipage}
\vspace{.1cm}

\noindent
Thus, $L$ and $R$ are the lift- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet they are not affected by the rule itself.
%
If a category is adhesive, and the arrows of the rules are monomorphisms, then the presence of a match ensures that the
two pushouts exist, hence a rewriting step can be performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{xxx} extend the core framework, still ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the benefiots of adhesivity
%the classical results from the theory of graph transformation 
can still be obtained~\cite{xxx}. 
If only the left-hand side belongs to $\mathcal{M}$, the current theory is stil under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being $\mathcal{M}$-adhesive can be a daunting task. For this reason, sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their adhesivity has not yet been settled.

>>>>

\vspace{.1cm}
\noindent
\begin{minipage}[l]{.78\linewidth}In short, following the double-pushout (DPO) approach
to graph transformation~\cite{ehrig1999handbook,ehrig2006fundamentals,corradini1997algebraic}, a rule is given by two arrows $l: K \to L$ and $r: K \rightarrow R$
and its execution requires a match $m: L \to G$ such that the rewriting step from $G$
to $H$ is given by the diagram aside, whose halves
  \end{minipage}%
    \hfill
  \begin{minipage}[r]{.20\linewidth }
    \xymatrix@C=.5cm@R=.5cm{
      L \ar[d]_{m}
      & K \ar[r]^r \ar[l]_{l} \ar[d] & R \ar [d] \\
      G & C \ar[r] \ar[l]                    & H
    }
%    \xymatrix@C=.5cm@R=.5cm{
%      L \ar[d]_{m} \ar@{}[dr]|{(1)}
%      & I \ar[r]^r \ar@{>->}[l]_{l} \ar[d] \ar@{}[dr]|{(2)} & R \ar [d] \\
%      G & C \ar@{->}[r] \ar@{>->}[l]                    & H\\
 %     & J \ar@{->}[ul] \ar@{->}[u]|{k} \ar@{->}[ur]
 %   }
  \end{minipage}
\vspace{.1cm}

\noindent
 are pushouts. Thus, $L$ and $R$ are the lift- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet they are not affected by the rule itself.
%
If a category is adhesive, and the arrows of the rules are monomorphisms, then the presence of a match ensures that the
two pushouts exist, hence a rewriting step can be performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{azzi2019essence,heindel2009category} extends the core framework, still ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the benefits of adhesivity
%the classical results from the theory of graph transformation 
can still be obtained~\cite{ehrig2012,ehrig2014adhesive}. 
If only the left-hand side belongs to $\mathcal{M}$, the current theory is stil under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being $\mathcal{M}$-adhesive can be a daunting task. For this reason, sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their $\mathcal{M}$-adhesivity has not yet been settled.
